---
title: "LMNA NMD prediction: benchmarking existing predictors against a gene-specific random forest"
subtitle: "Reproducible pipeline for Figure 7, Figure S7 and Table S7"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 4)
knitr::opts_chunk$set(cache.path = '.cache/', autodep = TRUE)
suppressPackageStartupMessages({
  library(readxl); library(ranger); library(mclust); library(ggplot2)
  library(gridExtra); library(grid); library(treeshap); library(scales)
  library(ggbeeswarm); library(openxlsx)
})
DATA_DIR <- "data"; OUT <- "output"; dir.create(OUT, showWarnings = FALSE)
SEED <- 42
COL_ACT <- "#3B7A6B"; COL_PRED <- "#C77B3B"; COL_TEST <- "#6A4C93"; COL_DISC <- "#E23B3B"; COL_EDIT <- "#8D6E9C"
# Journal geometry: 6.5 in text width, minimum 8 pt for every element,
# axis titles 10 pt, panel letters 12 pt bold.
theme_fig <- function(base = 8) theme_minimal(base_size = base) +
  theme(panel.grid = element_blank(), plot.title = element_text(face = "bold", size = 12, margin = margin(b = 2)),
        axis.title = element_text(size = 10), axis.text = element_text(size = 8),
        legend.text = element_text(size = 8, margin = margin(r = 8)), legend.title = element_text(size = 8),
        legend.spacing.x = unit(0.2, "cm"), plot.margin = margin(2, 3, 2, 3),
        strip.background = element_blank(), strip.text = element_text(size = 8),
        panel.border = element_rect(colour = "grey25", fill = NA, linewidth = 0.4),
        plot.title.position = "plot")
```

# Design

Everything below is computed in this document, in R, from `data/`.

1. **Held-out test set.** 10% of codons (all stop codons at a codon travel together)
   are drawn at random once, with a fixed seed, and are excluded from every
   training and model-selection step of that split: the Gaussian-mixture threshold, the
   codon-grouped cross-validation, the block-size sweep, the discordance threshold,
   the linear rescaling of NMDetective scores, and SHAP. The AUC and R² in Fig. 7A
   are medians over 50 independent re-draws of such a split (section 6a).
2. **Gene-agnostic features.** The random forest does not receive the hand-defined
   5′/active/3′ region, and is not told the shape of NMD activity along the CDS;
   the local shape and the positions of the transitions are learned. Positional features are derived
   from the NM_170707.4 exon structure (`data/NM_170707.4.gb`) in units that exist
   in every gene: nt from CDS start, nt to the last EJC, nt to the next EJC, nt from
   the previous EJC, containing-exon length, and number of downstream EJCs.
3. **Leakage-aware evaluation on the training set.** Cross-validation is grouped by
   codon (so a held-out PTC's sibling stop codons are never in training) and, for
   the block-size sweep, by contiguous windows of codons.
4. **The PTC is the unit of analysis.** Every count, threshold and plotted point in
   Figs. 7 and S7 is per PTC (n = 763), never a per-codon average. Codon means appear
   in exactly two places, both noted in the code: the misclassification comparison in
   section 11, kept only so the two units can be contrasted, and the boundary
   positions in Fig. 7A, where averaging is correct because a boundary is a property
   of position rather than of stop-codon identity.

# 1. Data and exon structure

```{r load}
lmna <- read_excel(file.path(DATA_DIR, "TableS6_NMD_efficiency_values.xlsx"))
lmna$mutation <- as.character(lmna$mutation)
lmna <- lmna[!is.na(lmna$mean_NMD_efficiency), ]
seq_lines <- readLines(file.path(DATA_DIR, "LMNA_NM_170707.4_CDS.fasta"))
cds_seq <- paste(seq_lines[!startsWith(seq_lines, ">")], collapse = "")
lmna$plus4 <- substring(cds_seq, (lmna$codon - 1) * 3 + 4, (lmna$codon - 1) * 3 + 4)
lmna$plus5 <- substring(cds_seq, (lmna$codon - 1) * 3 + 5, (lmna$codon - 1) * 3 + 5)
stopifnot(lmna$plus4[lmna$codon == 240][1] == "C", lmna$plus5[lmna$codon == 240][1] == "T")
nmdet <- read.csv(file.path(DATA_DIR, "NMDetective_LMNA_per_codon_scores.csv"))
n_before <- nrow(lmna)
lmna <- merge(lmna, nmdet, by = "codon"); lmna <- lmna[order(lmna$codon, lmna$mutation), ]
stopifnot(nrow(lmna) == n_before, nrow(lmna) == 763L)
rownames(lmna) <- NULL

gb <- readLines(file.path(DATA_DIR, "NM_170707.4.gb"))
ex <- regmatches(gb, regexec("^\\s+exon\\s+(\\d+)\\.\\.(\\d+)", gb))
exons <- do.call(rbind, lapply(ex[lengths(ex) > 0], function(m) as.integer(m[2:3])))
cds <- as.integer(regmatches(gb, regexec("^\\s+CDS\\s+(\\d+)\\.\\.(\\d+)", gb))[[grep("^\\s+CDS\\s", gb)]][2:3])
exon_cds <- exons - cds[1] + 1; exon_len <- exons[, 2] - exons[, 1] + 1
junctions <- exon_cds[-nrow(exon_cds), 2]; LAST_EJC <- tail(junctions, 1)
stopifnot(nrow(exons) == 12, LAST_EJC == 1968, nchar(cds_seq) == cds[2] - cds[1] + 1)
cat("PTC rows:", nrow(lmna), " codons:", length(unique(lmna$codon)),
    "\nExon ends (CDS codon):", round(junctions / 3, 1), "\n")

pos_features <- function(c) {
  nt_end <- 3 * c; ex_i <- which(exon_cds[, 1] <= nt_end & nt_end <= exon_cds[, 2])
  down <- junctions[junctions >= nt_end]
  c(nt_from_CDS_start = 3 * (c - 1), nt_to_last_EJC = LAST_EJC - nt_end,
    nt_to_next_EJC = if (length(down)) min(down) - nt_end else -1L,
    nt_from_prev_EJC = nt_end - (if (ex_i > 1) exon_cds[ex_i - 1, 2] else 0),
    exon_length = exon_len[ex_i], n_downstream_EJCs = length(down))
}
Fpos <- as.data.frame(t(sapply(lmna$codon, pos_features)))
onehot <- function(x, prefix) { lv <- sort(unique(x))
  out <- as.data.frame(lapply(lv[-1], function(l) as.integer(x == l))); names(out) <- paste0(prefix, "_", lv[-1]); out }
# Sequence context is encoded as three blocks:
#   (i) stop-codon identity, capturing the general TAA > TAG > TGA termination hierarchy;
#   (ii) the +4 and +5 nucleotides, one-hot encoded;
#   (iii) a single TGA-CT indicator. A cytosine at +4 with a uracil at +5 following a UGA is
#         the element required for translational readthrough, and its effect is conditional
#         on the stop codon rather than additive across +4 and +5, so the one-hot columns
#         alone cannot represent it. Adding the indicator to them improves classification
#         without cost to R2 (see Methods).
Fctx <- cbind(onehot(lmna$mutation, "stop"), onehot(lmna$plus4, "plus4"), onehot(lmna$plus5, "plus5"),
              TGACT = as.integer(lmna$mutation == "TGA" & lmna$plus4 == "C" & lmna$plus5 == "T"))
X_full <- cbind(Fpos, Fctx)
y <- lmna$mean_NMD_efficiency; codon <- lmna$codon; n <- nrow(lmna)
tile <- findInterval(codon, c(1, 214, 434, 537, 626))
```

# 2. Held-out split (10% of codons, random, grouped)

```{r split}
set.seed(SEED)
ucod <- sort(unique(codon)); n_test <- round(0.10 * length(ucod))
test_codons <- sort(sample(ucod, n_test))
is_test <- codon %in% test_codons; is_train <- !is_test
cat(sprintf("Held out %d codons (%d PTC rows); training on %d codons (%d rows)\n",
            n_test, sum(is_test), length(ucod) - n_test, sum(is_train)))
cat("Held-out codons:", test_codons, "\n")
```

# 3. Ground-truth labels: Gaussian-mixture threshold (training rows only)

```{r gmm}
gmm <- Mclust(y[is_train], G = 2, verbose = FALSE)
mu <- gmm$parameters$mean; o <- order(mu); mu <- mu[o]
sg <- sqrt(gmm$parameters$variance$sigmasq); if (length(sg) == 1) sg <- rep(sg, 2); sg <- sg[o]
w <- gmm$parameters$pro[o]
grid <- seq(-0.2, 1, length.out = 20000)
post <- w[2] * dnorm(grid, mu[2], sg[2]) / (w[1] * dnorm(grid, mu[1], sg[1]) + w[2] * dnorm(grid, mu[2], sg[2]))
THRESH <- grid[which(diff(sign(post - 0.5)) != 0)[1]]
lab <- as.integer(y > THRESH)
cat(sprintf("Components: mean %.3f/%.3f, sd %.3f/%.3f, weight %.2f/%.2f -> threshold %.3f\n", mu[1], mu[2], sg[1], sg[2], w[1], w[2], THRESH))
cat(sprintf("Held-out set: %d NMD-triggering, %d NMD-escaping rows\n", sum(lab[is_test]), sum(1 - lab[is_test])))
stopifnot(sum(lab[is_test]) > 0, sum(1 - lab[is_test]) > 0)
writeLines(as.character(THRESH), file.path(OUT, "gmm_threshold.txt"))
```

# 4. Comparator models

```{r comparators}
r2  <- function(a, p) 1 - sum((a - p)^2) / sum((a - mean(a))^2)
auc <- function(s, l) { r <- rank(s); n1 <- sum(l == 1); n0 <- sum(l == 0); (sum(r[l == 1]) - n1 * (n1 + 1) / 2) / (n1 * n0) }
# Rule boundaries are derived from the NM_170707.4 exon structure rather than hard-coded, and
# follow each tool's own implementation:
#   50-nt rule: a PTC escapes if it lies <= 50 nt upstream of the last EJC.
#   aenmd v0.3.11 (R/nmd_escape_rules.R; css_prox_dist = 150L, penultimate_prox_dist = 50L):
#     escape if start-proximal (3c - 3 < 150), in the last exon, within 50 nt of the penultimate
#     exon end, in a single-exon transcript, or in an exon of >= 407 nt. The long-exon rule is
#     evaluated on CDS-relative exon coordinates, where LMNA's largest coding exon is 356 nt, so
#     it applies to no exon of this transcript.
#   VEP NMD plugin (Ensembl/VEP_plugins, NMD.pm): escape if cds_end <= 101, in the last exon,
#     within the final 51 nt of the penultimate exon, or in an intronless transcript.
CSS_PROX <- 150L; PENULT_PROX_AENMD <- 50L; VEP_CDS_MAX <- 101L; VEP_PENULT <- 51L
cds_end_of <- function(cc) 3L * cc                       # 3'-most base of the codon
rule_5055  <- as.integer(LAST_EJC - cds_end_of(codon) > 50L)
rule_aenmd <- as.integer((3L * codon - 3L) >= CSS_PROX &
                         (LAST_EJC - cds_end_of(codon)) >= PENULT_PROX_AENMD)
rule_vep   <- as.integer(cds_end_of(codon) > VEP_CDS_MAX &
                         cds_end_of(codon) < LAST_EJC - VEP_PENULT)
stopifnot(identical(rule_5055,  as.integer(codon < 640)),
          identical(rule_aenmd, as.integer(codon > 50 & codon < 640)),
          identical(rule_vep,   as.integer(codon > 33 & codon < 639)))
# NMDetective: linear rescale to the measured scale, fit on TRAINING rows only
rescale_fit <- function(raw) lm(y[is_train] ~ raw[is_train])
fitA <- rescale_fit(lmna$NMDetectiveA); fitB <- rescale_fit(lmna$NMDetectiveB)
nmdA_scaled <- coef(fitA)[1] + coef(fitA)[2] * lmna$NMDetectiveA
nmdB_scaled <- coef(fitB)[1] + coef(fitB)[2] * lmna$NMDetectiveB
```

# 5. Random forest: training-set cross-validation

```{r rf-helpers}
fit_rf <- function(X, idx, importance = "none")
  ranger(y ~ ., data = cbind(X[idx, , drop = FALSE], y = y[idx]), num.trees = 500, min.node.size = 3,
         seed = SEED, importance = importance)
predict_rf <- function(m, X, idx) predict(m, data = X[idx, , drop = FALSE])$predictions
cv_oof <- function(X, folds, idx = which(is_train)) {
  oof <- rep(NA_real_, n)
  for (f in unique(folds[idx])) { te <- idx[folds[idx] == f]; tr <- idx[folds[idx] != f]
    oof[te] <- predict_rf(fit_rf(X, tr), X, te) }
  oof
}
set.seed(SEED)
tr_cod <- ucod[!ucod %in% test_codons]
fold_grouped <- setNames(sample(rep(1:10, length.out = length(tr_cod))), tr_cod)[as.character(codon)]
fold_random  <- rep(NA_integer_, n); fold_random[is_train] <- sample(rep(1:10, length.out = sum(is_train)))
# Blocked CV: codons are grouped into contiguous windows of b codons, and the windows
# are dealt round-robin into at most MAX_FOLDS folds. Each fold therefore still holds out
# contiguous b-codon windows (the quantity of interest), but small b no longer implies
# hundreds of refits; with few windows this reduces to leave-one-window-out.
MAX_FOLDS <- 20
block_folds <- function(b) {
  if (b == "tile") return(tile)
  w <- as.integer(factor((codon - 1) %/% as.integer(b)))
  nw <- length(unique(w))
  if (nw <= MAX_FOLDS) return(w)
  set.seed(SEED); map <- setNames(sample(rep(1:MAX_FOLDS, length.out = nw)), sort(unique(w)))
  as.integer(map[as.character(w)])
}
```

## 5a. Codon-grouped out-of-fold predictions

```{r grouped-cv}
tr <- which(is_train)
# Out-of-fold predictions under codon-grouped CV: a PTC's sibling stop codons are never
# in its training fold. These are the values plotted for training codons in Fig. 7C,D;
# held-out codons use the final model instead.
oof_grouped <- cv_oof(X_full, fold_grouped)
cat(sprintf("Codon-grouped CV on training rows: R2 %.3f, AUC %.3f\n",
            r2(y[tr], oof_grouped[tr]), auc(oof_grouped[tr], lab[tr])))
```

## 5b. Block-size sweep

```{r block-sweep, cache=TRUE}
idx_by_tr <- split(tr, codon[tr])
boot_ci <- function(stat, idx_by, B = 1000, seed = 1) { set.seed(seed)
  v <- replicate(B, { i <- unlist(idx_by[sample(names(idx_by), length(idx_by), replace = TRUE)], use.names = FALSE)
    tryCatch(stat(i), error = function(e) NA) }); quantile(v, c(.025, .975), na.rm = TRUE) }
# NMDetectiveA is a fixed genome-wide score: it is not fitted to LMNA and cannot learn from the
# training folds, so it enters this panel as a single reference level rather than as a curve. Its
# linear rescaling to the measured scale is estimated once on all training rows (two parameters
# over 687 PTCs, so the in-sample optimism is negligible and runs in the comparator's favour).
NMDA_REF <- local({ b <- coef(lm(y[tr] ~ lmna$NMDetectiveA[tr]))
                    r2(y[tr], b[1] + b[2] * lmna$NMDetectiveA[tr]) })
block_sizes <- c("1", "3", "5", "10", "20", "30", "50", "tile")
sweep <- do.call(rbind, lapply(block_sizes, function(b) {
  p  <- cv_oof(X_full, block_folds(b))
  ci <- boot_ci(function(i) r2(y[i], p[i]), idx_by_tr, B = 300)
  data.frame(block = b, R2 = r2(y[tr], p[tr]), R2_lo = ci[1], R2_hi = ci[2], AUC = auc(p[tr], lab[tr]))
}))
cat(sprintf("NMDetectiveA reference (fixed score, single rescale on training rows): R2 %.3f\n", NMDA_REF))
rownames(sweep) <- NULL; print(sweep, digits = 3, row.names = FALSE)
write.csv(sweep, file.path(OUT, "block_size_sweep.csv"), row.names = FALSE)
```

# 6. Final model

```{r final-model}
te <- which(is_test)
rf_full <- fit_rf(X_full, tr, importance = "permutation")
pred_full <- rep(NA_real_, n); pred_full[te] <- predict_rf(rf_full, X_full, te)
cat(sprintf("Final model fitted on %d training PTCs; predictions for %d held-out PTCs\n",
            length(tr), length(te)))
```

## 6a. Held-out evaluation over 50 random grouped splits

Every performance figure quoted in the paper comes from this section. The whole
procedure (split, rescale NMDetective on training rows, fit the random forest on
training rows, evaluate on held-out rows) is repeated over 50 random grouped
splits, and Fig. 7A reports the median and 5th-95th percentiles across them.

```{r repeated-splits, cache=TRUE}
N_SPLITS <- 50
set.seed(SEED + 1)
split_sets <- c(list(test_codons), replicate(N_SPLITS, sample(ucod, n_test), simplify = FALSE))
# element 1 is the prespecified split; elements 2..N_SPLITS+1 are the random re-draws
eval_split <- function(tc, k) {
  ite <- which(codon %in% tc); itr <- which(!codon %in% tc)
  fA <- lm(y[itr] ~ lmna$NMDetectiveA[itr]); sA <- coef(fA)[1] + coef(fA)[2] * lmna$NMDetectiveA
  fB <- lm(y[itr] ~ lmna$NMDetectiveB[itr]); sB <- coef(fB)[1] + coef(fB)[2] * lmna$NMDetectiveB
  pf <- predict_rf(fit_rf(X_full, itr), X_full, ite)
  sc <- list("50-55nt rule alone" = rule_5055[ite], "aenmd rules" = rule_aenmd[ite], "VEP predictor" = rule_vep[ite],
             "NMDetectiveB" = sB[ite], "NMDetectiveA (RF)" = sA[ite], "Our model" = pf)
  reg <- c(FALSE, FALSE, FALSE, TRUE, TRUE, TRUE)
  do.call(rbind, lapply(seq_along(sc), function(i) data.frame(split = k, model = names(sc)[i],
    AUC = auc(sc[[i]], lab[ite]), R2 = if (reg[i]) r2(y[ite], sc[[i]]) else NA_real_)))
}
rep_all <- do.call(rbind, lapply(seq_along(split_sets), function(i) eval_split(split_sets[[i]], i - 1L)))
prespec <- rep_all[rep_all$split == 0L, ]          # the prespecified split
rep_tab <- rep_all[rep_all$split > 0L, ]           # the 50 random re-draws
q <- function(v, p) as.numeric(quantile(v, p, na.rm = TRUE))
rep_sum <- do.call(rbind, lapply(split(rep_tab, rep_tab$model), function(d) data.frame(model = d$model[1],
  AUC_median = median(d$AUC), AUC_q05 = q(d$AUC, .05), AUC_q95 = q(d$AUC, .95),
  R2_median = median(d$R2), R2_q05 = q(d$R2, .05), R2_q95 = q(d$R2, .95),
  AUC_prespecified = prespec$AUC[prespec$model == d$model[1]],
  R2_prespecified  = prespec$R2[prespec$model == d$model[1]])))
rownames(rep_sum) <- NULL; print(rep_sum, digits = 3, row.names = FALSE)
write.csv(rep_tab, file.path(OUT, "repeated_splits.csv"), row.names = FALSE)
write.csv(rep_sum, file.path(OUT, "repeated_splits_summary.csv"), row.names = FALSE)
# percentile of the prespecified split within the 50 re-draws, per model
pct <- sapply(unique(rep_tab$model), function(m) { d <- rep_tab[rep_tab$model == m, ]
  c(AUC = mean(d$AUC <= prespec$AUC[prespec$model == m]), R2 = mean(d$R2 <= prespec$R2[prespec$model == m], na.rm = TRUE)) })
cat("\nPercentile of the prespecified split within the 50 re-draws:\n"); print(round(pct, 2))
```

## 6b. SHAP (exact TreeSHAP on the final model, training rows)

```{r shap}
unified <- ranger.unify(rf_full, X_full[tr, ])
shap_values <- treeshap(unified, X_full[tr, ], verbose = FALSE)$shaps
imp <- sort(colMeans(abs(shap_values)), decreasing = TRUE); print(round(imp, 4))
write.csv(data.frame(feature = names(imp), mean_abs_SHAP = imp, permutation_importance = importance(rf_full)[names(imp)]),
          file.path(OUT, "feature_importance.csv"), row.names = FALSE)
```

# 7. Per-codon table for plotting

Training codons use grouped-CV out-of-fold predictions; held-out codons use the final model's predictions.

```{r percodon}
pred_plot <- ifelse(is_test, pred_full, oof_grouped)
segments_list <- list(c(1, 118), c(214, 270), c(434, 460), c(537, 566), c(626, 656))
seg_levels <- sapply(segments_list, function(s) paste0(s[1], "-", s[2]))
seg_of <- function(cd) { s <- rep(NA_character_, length(cd)); for (sg in segments_list) s[cd >= sg[1] & cd <= sg[2]] <- paste0(sg[1], "-", sg[2]); factor(s, levels = seg_levels) }

# PTC-resolved frame. Individual PTCs, not codon averages, are the unit of analysis:
# averaging the three stop codons at a position hides exactly the stop-codon-specific
# effects this paper is about (codon 240 TGA measures 0.04 against a codon mean of 0.62).
ptc <- data.frame(codon = codon, stop = factor(lmna$mutation, levels = c("TAA", "TAG", "TGA")),
                  actual = y, ours = pred_plot, nmdA = nmdA_scaled, nmdB = nmdB_scaled,
                  test = is_test, label = lab, segment = seg_of(codon))
ptc$rule_5055 <- rule_5055; ptc$rule_aenmd <- rule_aenmd; ptc$rule_vep <- rule_vep

# Codon-averaged frame, retained only so the codon-level counts can be reported
# alongside the PTC-level ones for comparison.
pc <- aggregate(cbind(actual = y, ours = pred_plot, nmdA = nmdA_scaled, nmdB = nmdB_scaled) ~ codon, FUN = mean)
pc$test <- pc$codon %in% test_codons
pc$rule_5055  <- as.integer(LAST_EJC - 3L * pc$codon > 50L)
pc$rule_aenmd <- as.integer((3L * pc$codon - 3L) >= CSS_PROX & (LAST_EJC - 3L * pc$codon) >= PENULT_PROX_AENMD)
pc$rule_vep   <- as.integer(3L * pc$codon > VEP_CDS_MAX & 3L * pc$codon < LAST_EJC - VEP_PENULT)
pc$segment <- seg_of(pc$codon)

# Discordance threshold, now from PTC-resolved values in the active region (training only).
act_tr_ptc <- ptc$codon >= 25 & ptc$codon <= 635 & !ptc$test
act_tr_cod <- pc$codon  >= 25 & pc$codon  <= 635 & !pc$test
DISCORD       <- 2 * sd(ptc$actual[act_tr_ptc])
DISCORD_CODON <- 2 * sd(pc$actual[act_tr_cod])
cat(sprintf("Discordance threshold: %.3f per PTC | %.3f per codon mean\n",
            DISCORD, DISCORD_CODON))
cat(sprintf("Dataset: %d PTCs across %d codons; %s\n", nrow(ptc), length(unique(ptc$codon)),
            paste(sprintf("%d codons with %s PTCs", table(table(codon)), names(table(table(codon)))), collapse = ", ")))
strad <- sort(unique(codon[ave(lab, codon, FUN = function(v) length(unique(v))) > 1]))
cat(sprintf("Codons whose assayed stop codons straddle the threshold: %d (%s)\n",
            length(strad), paste(strad, collapse = ", ")))
cat(sprintf("PTCs whose label differs from their codon mean: %d of %d (%.1f%%)\n",
            sum(lab != as.integer(ave(y, codon, FUN = mean) > THRESH)), length(y),
            100 * mean(lab != as.integer(ave(y, codon, FUN = mean) > THRESH))))
```

# 7b. How many measured sites does a gene-specific model need?

If a gene-specific model requires saturation editing of the whole gene, it is of
little practical use. This asks the operational question instead: if you can assay
X PTC sites in a target gene, is it better to train a gene-specific model on them
or to keep using a genome-wide predictor? Both options are given the *same* X
sites, since those sites can equally be used to calibrate NMDetectiveA to the gene.
A fixed evaluation set (30% of codons) is drawn once and never trained on, so every
X is scored on identical codons; X codons are then drawn from the remaining 70%.

```{r learning-curve, cache=TRUE}
N_REP_LC <- 40; SIZES <- c(5, 10, 15, 20, 30, 40, 60, 80, 120, 160)
set.seed(SEED)
lc_eval <- sort(sample(ucod, round(0.30 * length(ucod))))
lc_pool <- setdiff(ucod, lc_eval); ie <- which(codon %in% lc_eval)
cat(sprintf("Learning-curve evaluation set: %d codons (%d PTCs), fixed; sampling pool %d codons\n",
            length(lc_eval), length(ie), length(lc_pool)))
# reference: NMDetectiveA calibrated on the entire sampling pool (a best case it
# could never beat by having more sites, since its ceiling is its own information)
f0 <- lm(y[codon %in% lc_pool] ~ lmna$NMDetectiveA[codon %in% lc_pool])
nmdA_pool <- coef(f0)[1] + coef(f0)[2] * lmna$NMDetectiveA
LC_REF_R2 <- r2(y[ie], nmdA_pool[ie]); LC_REF_AUC <- auc(nmdA_pool[ie], lab[ie])
X_hybrid <- cbind(X_full, nmdA = lmna$NMDetectiveA)
lc_rows <- list()
for (X in SIZES) for (rep in 1:N_REP_LC) {
  set.seed(SEED * 1000 + X * 10 + rep)
  # one PTC per sampled codon: a targeted experiment installs a single stop per position
  tc <- sample(lc_pool, X)
  it <- unlist(lapply(tc, function(cc) { i <- which(codon == cc); i[sample(length(i), 1)] }))
  fit <- try(lm(y[it] ~ lmna$NMDetectiveA[it]), silent = TRUE)
  p_cal <- if (inherits(fit, "try-error") || any(is.na(coef(fit)))) rep(mean(y[it]), length(ie)) else
    coef(fit)[1] + coef(fit)[2] * lmna$NMDetectiveA[ie]
  p_rf <- predict_rf(fit_rf(X_full, it), X_full, ie)
  p_hy <- predict(ranger(y ~ ., data = cbind(X_hybrid[it, ], y = y[it]), num.trees = 500,
                         min.node.size = 3, seed = SEED), data = X_hybrid[ie, ])$predictions
  lc_rows[[length(lc_rows) + 1]] <- data.frame(X = X, rep = rep,
    model = c("NMDetectiveA (calibrated on X sites)", "Gene-specific RF", "Hybrid RF + NMDetectiveA"),
    R2 = c(r2(y[ie], p_cal), r2(y[ie], p_rf), r2(y[ie], p_hy)),
    AUC = c(auc(p_cal, lab[ie]), auc(p_rf, lab[ie]), auc(p_hy, lab[ie])))
}
lc_raw <- do.call(rbind, lc_rows)
lc_raw$n_PTCs <- lc_raw$X   # one PTC per sampled codon
qq <- function(v, p) as.numeric(quantile(v, p, na.rm = TRUE))
lc <- do.call(rbind, lapply(split(lc_raw, list(lc_raw$X, lc_raw$model), drop = TRUE), function(d)
  data.frame(X = d$X[1], n_PTCs = d$n_PTCs[1], model = d$model[1],
             R2_median = median(d$R2, na.rm = TRUE), R2_q05 = qq(d$R2, .05), R2_q95 = qq(d$R2, .95),
             AUC_median = median(d$AUC, na.rm = TRUE), AUC_q05 = qq(d$AUC, .05), AUC_q95 = qq(d$AUC, .95),
             frac_beat_ref_R2 = mean(d$R2 > LC_REF_R2, na.rm = TRUE))))
lc <- lc[order(lc$model, lc$X), ]; rownames(lc) <- NULL
lc$ref_R2 <- LC_REF_R2; lc$ref_AUC <- LC_REF_AUC
print(lc, digits = 3, row.names = FALSE)
write.csv(lc, file.path(OUT, "learning_curve.csv"), row.names = FALSE)
write.csv(lc_raw, file.path(OUT, "learning_curve_raw.csv"), row.names = FALSE)
gsel <- lc[lc$model == "Gene-specific RF", ]
X_CROSS <- gsel$X[which(gsel$R2_median > LC_REF_R2)[1]]
X_ROBUST <- gsel$X[which(gsel$frac_beat_ref_R2 >= 0.99)[1]]
cat(sprintf("\nReference (NMDetectiveA, whole-pool calibration): R2 %.3f, AUC %.3f\n", LC_REF_R2, LC_REF_AUC))
cat(sprintf("Gene-specific RF median exceeds it from X = %d codons (~%d PTCs); exceeds it in >=99%% of draws from X = %d (~%d PTCs)\n",
            X_CROSS, gsel$n_PTCs[gsel$X == X_CROSS], X_ROBUST, gsel$n_PTCs[gsel$X == X_ROBUST]))
lc_plot <- function(metric, med, lo, hi, ref, ttl, ylab, ylim, legend = FALSE) {
  d <- lc; d$model <- factor(d$model,
    levels = c("NMDetectiveA (calibrated on X sites)", "Gene-specific RF", "Hybrid RF + NMDetectiveA"),
    labels = c("NMDetectiveA, calibrated on the X sites", "Gene-specific RF", "Hybrid (RF + NMDetectiveA feature)"))
  ggplot(d, aes(X, .data[[med]], color = model, fill = model)) +
    geom_hline(yintercept = ref, linetype = "dashed", color = "grey35", linewidth = .4) +
    geom_ribbon(aes(ymin = .data[[lo]], ymax = .data[[hi]]), alpha = .15, color = NA) +
    geom_line(linewidth = .7) + geom_point(size = 1.3) +
    scale_color_manual(values = c("grey55", COL_ACT, COL_PRED)) +
    scale_fill_manual(values = c("grey55", COL_ACT, COL_PRED)) +
    scale_x_log10(breaks = c(5, 10, 20, 40, 80, 160)) + coord_cartesian(ylim = ylim) +
    labs(title = ttl, x = "Codons sampled (1 PTC each)", y = ylab) + theme_fig() +
    guides(color = guide_legend(ncol = 1), fill = guide_legend(ncol = 1)) +
    theme(panel.grid.minor = element_blank(), axis.title = element_text(size = 8),
          legend.position = if (legend) "bottom" else "none", legend.title = element_blank(),
          legend.text = element_text(size = 8), legend.key.size = unit(.35, "lines"),
          legend.box.margin = margin(-4, 0, 0, 0))
}
```

## Sampling design: which codons a base edit can reach

```{r sampling-design, cache=TRUE}
# A stop codon can only be installed by a single nucleotide edit if it lies one
# substitution from the reference codon. In LMNA only 76 of 263 assayed codons
# have any such stop, and they under-sample the 3' boundary badly, so this
# constrains both how many and which positions a base-editing experiment reaches.
lmna$ref_codon <- substring(cds_seq, (lmna$codon - 1) * 3 + 1, (lmna$codon - 1) * 3 + 3)
lmna$n_edits <- mapply(function(r, m) sum(strsplit(r, "")[[1]] != strsplit(m, "")[[1]]),
                       lmna$ref_codon, lmna$mutation)
one_edit <- lmna$n_edits == 1
pool_edit <- intersect(lc_pool, unique(codon[one_edit]))
cat(sprintf("Edit-reachable: %d of %d assayed codons; %d of %d in the sampling pool\n",
            length(unique(codon[one_edit])), length(ucod), length(pool_edit), length(lc_pool)))
tb <- function(v) table(cut(v, c(0, 118, 270, 460, 566, 656),
                            labels = c("1-118", "214-270", "434-460", "537-566", "626-656")))
print(rbind(full_pool = tb(lc_pool), edit_reachable = tb(pool_edit)))

sd_rows <- list()
for (X in SIZES) for (rep in 1:N_REP_LC) {
  set.seed(SEED * 1000 + X * 10 + rep)
  tc <- sample(lc_pool, X)
  idx_one <- unlist(lapply(tc, function(cc) { i <- which(codon == cc); i[sample(length(i), 1)] }))
  p_rf <- predict_rf(fit_rf(X_full, idx_one), X_full, ie)
  sd_rows[[length(sd_rows) + 1]] <- data.frame(X = X, rep = rep, mode = "Any codon",
    n_PTCs = length(idx_one), R2 = r2(y[ie], p_rf), AUC = auc(p_rf, lab[ie]))
}
# The edit-reachable pool is smaller than the full pool, so it gets its own size grid
# running up to the largest round number the pool supports rather than stopping at 40.
SIZES_EDIT <- c(SIZES[SIZES <= length(pool_edit)], 50)
SIZES_EDIT <- sort(unique(SIZES_EDIT[SIZES_EDIT <= length(pool_edit)]))
cat("Edit-reachable training sizes:", SIZES_EDIT, "(pool =", length(pool_edit), "codons)\n")
for (X in SIZES_EDIT) for (rep in 1:N_REP_LC) {
  set.seed(SEED * 2000 + X * 10 + rep)
  tce <- sample(pool_edit, X)
  ite <- unlist(lapply(tce, function(cc) { i <- which(codon == cc & one_edit); i[sample(length(i), 1)] }))
  p_rf <- predict_rf(fit_rf(X_full, ite), X_full, ie)
  sd_rows[[length(sd_rows) + 1]] <- data.frame(X = X, rep = rep, mode = "Single-edit-reachable codons only",
    n_PTCs = length(ite), R2 = r2(y[ie], p_rf), AUC = auc(p_rf, lab[ie]))
}
sd_raw <- do.call(rbind, sd_rows)
sd <- do.call(rbind, lapply(split(sd_raw, list(sd_raw$X, sd_raw$mode), drop = TRUE), function(d)
  data.frame(X = d$X[1], mode = d$mode[1], n_PTCs = round(mean(d$n_PTCs)),
             R2_median = median(d$R2, na.rm = TRUE), R2_q05 = qq(d$R2, .05), R2_q95 = qq(d$R2, .95),
             AUC_median = median(d$AUC, na.rm = TRUE), frac_beat = mean(d$R2 > LC_REF_R2, na.rm = TRUE))))
rownames(sd) <- NULL; sd <- sd[order(sd$mode, sd$X), ]
print(sd, digits = 3, row.names = FALSE)
write.csv(sd, file.path(OUT, "sampling_design.csv"), row.names = FALSE)
write.csv(sd_raw, file.path(OUT, "sampling_design_raw.csv"), row.names = FALSE)
for (mm in unique(sd$mode)) { h <- sd[sd$mode == mm, ]; i <- which(h$R2_median > LC_REF_R2)[1]
  cat(sprintf("Crossover vs NMDetectiveA, %-36s %s\n", mm,
              if (is.na(i)) "not reached" else sprintf("X = %d codons (%d PTCs)", h$X[i], h$n_PTCs[i]))) }
s1 <- sd[sd$mode == "Any codon", ]; sed <- sd[sd$mode == "Single-edit-reachable codons only", ]
fmt_at <- function(h, N) if (N %in% h$X) sprintf("%.3f", h$R2_median[h$X == N]) else "     -"
for (N in sort(union(s1$X, sed$X)))
  cat(sprintf("  %3d codons: any codon R2 %s | edit-reachable R2 %s\n", N, fmt_at(s1, N), fmt_at(sed, N)))
```

# 9. Independent gene: BRCA1 (Findlay et al., 2018)

The readout there is relative RNA abundance, not SMG1i-based NMD efficiency, and only
codons in two disjoint blocks were assayed, so the attainable R2 is lower; the shapes of
the three curves are the comparable quantity.

```{r brca1, cache=TRUE}
Btx <- local({ gb <- readLines(file.path(DATA_DIR, "NM_007294.4.gb"))
  e <- regmatches(gb, regexec("^\\s+exon\\s+(\\d+)\\.\\.(\\d+)", gb))
  ex <- do.call(rbind, lapply(e[lengths(e) > 0], function(m) as.integer(m[2:3])))
  cd <- as.integer(regmatches(gb, regexec("^\\s+CDS\\s+(\\d+)\\.\\.(\\d+)", gb))[[grep("^\\s+CDS\\s", gb)]][2:3])
  list(ec = ex - cd[1] + 1, el = ex[, 2] - ex[, 1] + 1, ncod = (cd[2] - cd[1] + 1) / 3 - 1) })
bfeat <- function(c) { J <- Btx$ec[-nrow(Btx$ec), 2]; L <- tail(J, 1); nt <- 3 * c
  e <- which(Btx$ec[, 1] <= nt & nt <= Btx$ec[, 2]); dn <- J[J >= nt]
  c(nt_from_CDS_start = 3 * (c - 1), nt_to_last_EJC = L - nt,
    nt_to_next_EJC = if (length(dn)) min(dn) - nt else -1L,
    nt_from_prev_EJC = nt - (if (e > 1) Btx$ec[e - 1, 2] else 0),
    exon_length = Btx$el[e], n_downstream_EJCs = length(dn)) }
bb <- read_excel(file.path(DATA_DIR, "Findlay2018_BRCA1_SupplementaryTable.xlsx"), skip = 2)
bb$mean.rna.score <- suppressWarnings(as.numeric(bb$mean.rna.score))
bb <- bb[bb$consequence == "Nonsense" & !is.na(bb$mean.rna.score), ]
bb$codon <- ceiling(suppressWarnings(as.integer(bb$transcript_position)) / 3)
bb <- bb[!is.na(bb$codon) & bb$codon >= 1 & bb$codon <= Btx$ncod - 1, ]
bnm <- read.csv(file.path(DATA_DIR, "NMDetective_BRCA1_per_codon.csv"))
bb$nmdA <- bnm$NMDetectiveA[match(bb$codon, bnm$codon)]
bb <- bb[!is.na(bb$nmdA), ]
bF <- as.data.frame(t(sapply(bb$codon, bfeat)))
by <- as.numeric(bb$mean.rna.score); bc <- bb$codon; buc <- sort(unique(bc))
set.seed(SEED); bev <- sort(sample(buc, round(0.30 * length(buc)))); bpool <- setdiff(buc, bev)
bie <- which(bc %in% bev)
cat(sprintf("BRCA1: %d PTCs, %d codons; eval %d codons (%d PTCs), pool %d codons\n",
            nrow(bb), length(buc), length(bev), length(bie), length(bpool)))
bsz <- c(10, 20, 30, 50); brow <- list()
for (X in bsz) for (rep in 1:N_REP_LC) {
  set.seed(SEED * 3000 + X * 10 + rep)
  tc <- sample(bpool, X)
  it <- unlist(lapply(tc, function(cc) { i <- which(bc == cc); i[sample(length(i), 1)] }))
  ff <- function(M) predict(ranger(y ~ ., data = cbind(M[it, , drop = FALSE], y = by[it]),
                                   num.trees = 500, min.node.size = 3, seed = SEED),
                            data = M[bie, , drop = FALSE])$predictions
  fit <- try(lm(by[it] ~ bb$nmdA[it]), silent = TRUE)
  pc2 <- if (inherits(fit, "try-error") || any(is.na(coef(fit)))) rep(mean(by[it]), length(bie)) else
    coef(fit)[1] + coef(fit)[2] * bb$nmdA[bie]
  brow[[length(brow) + 1]] <- rbind(
    data.frame(X = X, rep = rep, model = "NMDetectiveA, calibrated", R2 = r2(by[bie], pc2)),
    data.frame(X = X, rep = rep, model = "Gene-specific model", R2 = r2(by[bie], ff(bF))),
    data.frame(X = X, rep = rep, model = "Gene-specific + NMDetectiveA", R2 = r2(by[bie], ff(cbind(bF, nmdA = bb$nmdA)))))
}
braw <- do.call(rbind, brow)
bsum <- do.call(rbind, lapply(split(braw, list(braw$X, braw$model), drop = TRUE), function(k)
  data.frame(X = k$X[1], model = k$model[1], R2_median = median(k$R2),
             R2_q05 = qq(k$R2, .05), R2_q95 = qq(k$R2, .95))))
rownames(bsum) <- NULL; print(bsum, digits = 3, row.names = FALSE)
write.csv(bsum, file.path(OUT, "BRCA1_replication.csv"), row.names = FALSE)
```

# 10. Boundary positions on the full transcript

Where each model's prediction crosses the classification threshold, computed across all
664 codons so the gaps between assayed regions cannot distort it. For the learned models
this is a measured property of the output, **not** a rule inside the model.

```{r boundaries}
allc <- nmdet[order(nmdet$codon), ]
allc$nmdA <- coef(fitA)[1] + coef(fitA)[2] * allc$NMDetectiveA
allc$nmdB <- coef(fitB)[1] + coef(fitB)[2] * allc$NMDetectiveB
grid_rows <- do.call(rbind, lapply(c("TAA", "TAG", "TGA"), function(st) {
  g <- as.data.frame(t(sapply(allc$codon, pos_features)))
  g$mutation <- st
  g$plus4 <- substring(cds_seq, (allc$codon - 1) * 3 + 4, (allc$codon - 1) * 3 + 4)
  g$plus5 <- substring(cds_seq, (allc$codon - 1) * 3 + 5, (allc$codon - 1) * 3 + 5)
  g$codon <- allc$codon; g }))
grid_rows <- grid_rows[grid_rows$plus4 != "" & grid_rows$plus5 != "", ]
Xg <- cbind(grid_rows[, names(Fpos)], onehot(grid_rows$mutation, "stop"),
            onehot(grid_rows$plus4, "plus4"), onehot(grid_rows$plus5, "plus5"))
# The TGA-CT indicator is set from the grid's own sequence context.
Xg$TGACT <- as.integer(grid_rows$mutation == "TGA" & grid_rows$plus4 == "C" & grid_rows$plus5 == "T")
for (cn in setdiff(names(X_full), names(Xg))) Xg[[cn]] <- 0
Xg <- Xg[, names(X_full)]
grid_rows$pred <- predict(rf_full, data = Xg)$predictions
allc$ours <- tapply(grid_rows$pred, grid_rows$codon, mean)[as.character(allc$codon)]

# CONVENTION. cross_up reports nt from the CDS start to the last codon that still escapes;
# cross_down reports nt from the last EJC to the last codon that still triggers. The
# rule-based rows are read off their own 0/1 vectors and land one codon later at the 5'
# end (e.g. VEP prints 99 nt for a rule whose cut is at cds_end 101). Both are reported as published.
cross_up <- function(v, cd = allc$codon) { s <- cd <= 200; cc <- cd[s]
  i <- which(diff(sign(v[s] - THRESH)) != 0)[1]; if (is.na(i)) NA else 3 * (cc[i] - 1) }
cross_down <- function(v, cd = allc$codon) { s <- cd >= 300; cc <- cd[s]
  i <- which(diff(sign(v[s] - THRESH)) != 0); if (!length(i)) NA else LAST_EJC - 3 * cc[tail(i, 1)] }
sens_cross <- function(v) { r <- sapply(c(.50, .55, .602, .65, .70), function(t) {
    s <- allc$codon >= 300; cc <- allc$codon[s]
    i <- which(diff(sign(v[s] - t)) != 0); if (!length(i)) NA else LAST_EJC - 3 * cc[tail(i, 1)] })
  range(r, na.rm = TRUE) }
rule_bounds <- function(v) { up <- which(diff(v) == 1)[1]; dn <- which(diff(v) == -1)[1]
  sprintf("%s / %s", if (is.na(up)) "-" else 3 * allc$codon[up],
                     if (is.na(dn)) "-" else LAST_EJC - 3 * allc$codon[dn]) }
emp <- function(v, star = FALSE, cd = allc$codon) sprintf("%d / %d%s", cross_up(v, cd), cross_down(v, cd), if (star) "*" else "")
emp_5055  <- rule_bounds(as.integer(LAST_EJC - 3L * allc$codon > 50L))
emp_aenmd <- rule_bounds(as.integer((3L * allc$codon - 3L) >= CSS_PROX &
                                    (LAST_EJC - 3L * allc$codon) >= PENULT_PROX_AENMD))
emp_vep   <- rule_bounds(as.integer(3L * allc$codon > VEP_CDS_MAX &
                                    3L * allc$codon < LAST_EJC - VEP_PENULT))
emp_A <- emp(allc$nmdA, TRUE); emp_B <- emp(allc$nmdB, TRUE)
emp_ours <- emp(allc$ours, TRUE); emp_act <- emp(pc$actual, FALSE, pc$codon)
sA <- sens_cross(allc$nmdA); sB <- sens_cross(allc$nmdB); sO <- sens_cross(allc$ours)
cat(sprintf("Boundaries (5' nt from CDS start / 3' nt upstream of last EJC)\n  measured %s | NMDetectiveA %s | NMDetectiveB %s | ours %s\n  rules: 50-nt %s | aenmd %s | VEP %s\n",
            emp_act, emp_A, emp_B, emp_ours, emp_5055, emp_aenmd, emp_vep))
cat(sprintf("3' crossover across thresholds 0.50-0.70 (nt to EJC): NMDetectiveA %d-%d | NMDetectiveB %d-%d | ours %d-%d\n",
            sA[1], sA[2], sB[1], sB[2], sO[1], sO[2]))
```

# 11. Figure 7

```{r figure7, fig.width=6.5, fig.height=8.75}
STOPCOL <- c(TAA = COL_ACT, TAG = COL_TEST, TGA = COL_PRED)
STOPSHP <- c(TAA = 21, TAG = 24, TGA = 22)

# Misclassification is counted per PTC. The codon-averaged counts are kept alongside
# only for comparison; averaging moves codon 240 (TGA 0.04, mean 0.62) across the
# threshold and scores every model as correct there.
err_ptc <- list("50-nt rule" = ptc$rule_5055, "aenmd" = ptc$rule_aenmd, "VEP" = ptc$rule_vep,
                "NMDetectiveB" = as.integer(ptc$nmdB > THRESH),
                "NMDetectiveA" = as.integer(ptc$nmdA > THRESH),
                "Our model"    = as.integer(ptc$ours > THRESH))
n_ptc <- sapply(err_ptc, function(p) sum(p != ptc$label))
true_c <- as.integer(pc$actual > THRESH)
err_cod <- list("50-nt rule" = pc$rule_5055, "aenmd" = pc$rule_aenmd, "VEP" = pc$rule_vep,
                "NMDetectiveB" = as.integer(pc$nmdB > THRESH),
                "NMDetectiveA" = as.integer(pc$nmdA > THRESH),
                "Our model"    = as.integer(pc$ours > THRESH))
n_cod <- sapply(err_cod, function(p) sum(p != true_c))
mis <- data.frame(Model = names(n_ptc),
                  codons_wrong = as.integer(n_cod), pct_codons = round(100 * n_cod / nrow(pc), 1),
                  PTCs_wrong  = as.integer(n_ptc), pct_PTCs   = round(100 * n_ptc / nrow(ptc), 1))
print(mis, row.names = FALSE)
write.csv(mis, file.path(OUT, "misclassification_by_unit.csv"), row.names = FALSE)

## Panel A -------------------------------------------------------------------
fmt <- function(v, lo, hi) ifelse(is.na(v), "-", sprintf("%.2f\n[%.2f-%.2f]", v, lo, hi))
key <- c("50-55nt rule alone", "aenmd rules", "VEP predictor", "NMDetectiveB", "NMDetectiveA (RF)", "Our model")
rows <- rbind(
 c("50-nt rule",   "Rules-based",   "Imposed",        emp_5055,  n_ptc[["50-nt rule"]]),
 c("aenmd",        "Rules-based",   "Imposed",        emp_aenmd, n_ptc[["aenmd"]]),
 c("VEP",          "Rules-based",   "Imposed",        emp_vep,   n_ptc[["VEP"]]),
 c("NMDetectiveB", "Decision tree", "Learned (TCGA)", emp_B,     n_ptc[["NMDetectiveB"]]),
 c("NMDetectiveA", "Random forest", "Learned (TCGA)", emp_A,     n_ptc[["NMDetectiveA"]]),
 c("Our model",    "Random forest", "Learned (LMNA)", emp_ours,  n_ptc[["Our model"]]))
rs <- rep_sum; rownames(rs) <- rs$model
tabA <- data.frame(rows,
  fmt(rs[key, "AUC_median"], rs[key, "AUC_q05"], rs[key, "AUC_q95"]),
  fmt(rs[key, "R2_median"],  rs[key, "R2_q05"],  rs[key, "R2_q95"]), stringsAsFactors = FALSE)
colnames(tabA) <- c("Model", "Type", "Positional\nboundaries", "Boundary on LMNA\n(start / EJC, nt)",
                    "PTCs\nmisclassified\n(FP + FN, of 763)", "AUC\nmed. [5-95%]", "R2\nmed. [5-95%]")
nr <- nrow(tabA); nc <- ncol(tabA); ours <- nr
fill <- matrix("white", nr, nc); fill[seq(1, nr, 2), ] <- "#F5F5F5"; fill[ours, ] <- COL_ACT
col  <- matrix("#1A1A1A", nr, nc); col[ours, ] <- "white"
face <- matrix("plain", nr, nc); face[, c(1, nc - 2, nc - 1, nc)] <- "bold"
fs   <- matrix(6, nr, nc); fs[, 1] <- 8
tt <- ttheme_minimal(base_size = 6,
  core = list(bg_params = list(fill = fill, col = "white", lwd = 1),
              fg_params = list(col = col, fontface = face, fontsize = as.vector(fs),
                               hjust = c(rep(0, nr), rep(.5, nr * (nc - 1))),
                               x = c(rep(.03, nr), rep(.5, nr * (nc - 1))))),
  colhead = list(bg_params = list(fill = "#D8D8D8", col = "white", lwd = 1),
                 fg_params = list(fontface = "bold", fontsize = c(8, rep(6, nc - 1)))))
tgA <- tableGrob(tabA, rows = NULL, theme = tt)
tgA$widths  <- unit(c(2.40, 2.00, 2.10, 2.40, 2.20, 1.70, 1.70), "cm")
tgA$heights <- unit(c(1.00, rep(.64, nr)), "cm")
panel_A <- arrangeGrob(tgA, top = textGrob("A", x = unit(.01, "npc"), hjust = 0,
                                           gp = gpar(fontface = "bold", fontsize = 12)))

## Panel B : comparators collapsed (position-only, so identical across stops) --
# The five comparators score a codon, not a stop codon: their prediction is the same
# for TAA/TAG/TGA at every one of the 263 codons. Their three rows would therefore
# differ only where the *measured* label differs between stops (21 straddling codons),
# so each is drawn as a single row, marked as an error if any PTC at that codon is
# misclassified. No codon carries both an FP and an FN for any comparator, so the
# collapsed row is unambiguous. Only our model, which can resolve stop identity,
# keeps three rows.
COMPARATORS <- setdiff(names(err_ptc), "Our model")
err_long <- do.call(rbind, lapply(names(err_ptc), function(m) {
  d <- data.frame(codon = ptc$codon, segment = ptc$segment, stop = as.character(ptc$stop),
                  model = m, class = ifelse(err_ptc[[m]] == ptc$label, "Correct",
                                            ifelse(err_ptc[[m]] == 1, "FP", "FN")),
                  stringsAsFactors = FALSE)
  if (m %in% COMPARATORS) {
    stopifnot(!any(tapply(d$class, d$codon, function(v) all(c("FP", "FN") %in% v))))
    d <- do.call(rbind, lapply(split(d, d$codon), function(g) {
      g$stop <- "All"
      g$class <- if (any(g$class == "FP")) "FP" else if (any(g$class == "FN")) "FN" else "Correct"
      g[1, , drop = FALSE] }))
  }
  d }))
# Held-out codons are a property of the data split, not of any one model, so the
# marking appears once as its own track rather than repeated on all six rows.
ho <- unique(ptc[ptc$test & !is.na(ptc$segment), c("codon", "segment")])
ho$stop <- ""; ho$model <- "Held out"; ho$class <- "Heldout"   # not a stop-codon category
err_long <- rbind(err_long, ho[, names(err_long)])
err_long$model <- factor(err_long$model, levels = c(names(err_ptc), "Held out"))
err_long$stop  <- factor(err_long$stop, levels = c("", "All", "TGA", "TAG", "TAA"))
cat(sprintf("Held-out codons marked in panel B: %d\n", nrow(ho)))
LAB <- c(Correct = "Correct", FP = "False positive (predicted triggering, escapes)",
         FN = "False negative (predicted escape, triggers)", Heldout = "Held-out codon")
make_panel_B <- function(border = TRUE) {
  g <- ggplot(err_long, aes(codon, stop, color = class, shape = class)) +
    geom_point(size = 1.15) +
    scale_color_manual(values = c(Correct = "#D9D9D9", FP = "#E0733A",
                                  FN = "#4A6FA5", Heldout = COL_TEST),
                       breaks = names(LAB), labels = LAB, name = NULL) +
    scale_shape_manual(values = c(Correct = 15, FP = 15, FN = 15, Heldout = 17),
                       breaks = names(LAB), labels = LAB, name = NULL) +
    facet_grid(model ~ segment, scales = "free", space = "free", switch = "y") +
    scale_x_continuous(breaks = c(0, 60, 240, 450, 550, 640)) +
    labs(title = "B", x = "Codon position in CDS", y = NULL) + theme_fig() +
    guides(color = guide_legend(nrow = 1, override.aes = list(size = 1.6)),
           shape = guide_legend(nrow = 1)) +
    theme(legend.position = "top", legend.key.size = unit(.35, "lines"),
          strip.text.x = element_blank(), strip.placement = "outside",
          strip.text.y.left = element_text(angle = 0, hjust = 1, size = 7),
          legend.margin = margin(0, 0, 0, 0), legend.box.margin = margin(0, 0, -4, 0),
          legend.text = element_text(size = 6.5, margin = margin(r = 4)),
          axis.text.y = element_text(size = 6.5), panel.spacing.y = unit(.14, "lines"),
          plot.margin = margin(8, 3, 2, 6))
  if (!border) g <- g + theme(panel.border = element_blank(),
                              axis.line.x = element_line(colour = "grey25", linewidth = .3))
  g
}
panel_B    <- make_panel_B(TRUE)
panel_B_nb <- make_panel_B(FALSE)

## Panels C and D : every PTC, not codon means -------------------------------
ptc_panel <- function(predcol, letter, show_leg) {
  x <- ptc[!is.na(ptc$segment), ]; x$pred <- x[[predcol]]
  # NMDetectiveA/B scores are published per nucleotide position, so their three
  # curves coincide; draw one neutral line in that case rather than three overplotted.
  varies <- any(tapply(x$pred, x$codon, function(v) length(unique(round(v, 9)))) > 1)
  x$disc <- abs(x$actual - x$pred) > DISCORD
  dc <- unique(x[x$disc, c("codon", "segment")])
  ggplot(x, aes(codon)) +
    { if (nrow(dc)) geom_rect(data = dc, aes(xmin = codon - .5, xmax = codon + .5, ymin = -Inf, ymax = Inf),
                              fill = COL_DISC, alpha = .15, inherit.aes = FALSE) } +
    # Measured PTCs are drawn first and semi-transparent so that the prediction
    # line, drawn last over a white halo, stays readable through the dense regions.
    geom_point(aes(y = actual, colour = stop, shape = stop,
                   fill = ifelse(test, "white", NA)), size = 1.15, stroke = .4, alpha = .5) +
    geom_line(aes(y = pred, group = stop), colour = "white", linewidth = 1.25, alpha = .9) +
    { if (varies) geom_line(aes(y = pred, colour = stop, group = stop), linewidth = .55)
      else geom_line(aes(y = pred, group = stop), colour = "grey25", linewidth = .6) } +
    scale_colour_manual(values = STOPCOL, name = NULL) +
    scale_shape_manual(values = STOPSHP, name = NULL) +
    scale_fill_identity(guide = "none") +
    facet_grid(~segment, scales = "free_x", space = "free_x") +
    coord_cartesian(ylim = c(-.12, 1.05)) +
    scale_x_continuous(breaks = c(0, 60, 240, 450, 550, 640)) +
    labs(title = letter, x = "Codon position in CDS", y = "NMD efficiency") + theme_fig() +
    guides(colour = guide_legend(nrow = 1, override.aes = list(linetype = 0, size = 1.8, alpha = 1)),
           shape = guide_legend(nrow = 1)) +
    theme(legend.position = if (show_leg) "top" else "none",
          legend.key.size = unit(.4, "lines"), legend.margin = margin(0, 0, 0, 0),
          legend.box.margin = margin(0, 0, -5, 0), legend.text = element_text(size = 7, margin = margin(r = 6)),
          panel.spacing = unit(.4, "lines"))
}
panel_C <- ptc_panel("nmdA", "C", TRUE)
panel_D <- ptc_panel("ours", "D", FALSE)
cat(sprintf("Discordant PTCs: ours %d/%d, NMDetectiveA %d/%d (codons with >=1: %d and %d)\n",
            sum(abs(ptc$actual - ptc$ours) > DISCORD), nrow(ptc),
            sum(abs(ptc$actual - ptc$nmdA) > DISCORD), nrow(ptc),
            length(unique(ptc$codon[abs(ptc$actual - ptc$ours) > DISCORD])),
            length(unique(ptc$codon[abs(ptc$actual - ptc$nmdA) > DISCORD]))))

## Panel E -------------------------------------------------------------------
lcp <- lc
lcp$model <- factor(lcp$model, levels = c("NMDetectiveA (calibrated on X sites)", "Gene-specific RF",
                                          "Hybrid RF + NMDetectiveA"),
  labels = c("NMDetectiveA, calibrated", "Gene-specific model", "Hybrid model"))
panel_E <- ggplot(lcp, aes(X, R2_median, color = model, fill = model)) +
  geom_hline(yintercept = lcp$ref_R2[1], linetype = "dashed", color = "grey35", linewidth = .4) +
  geom_ribbon(aes(ymin = R2_q05, ymax = R2_q95), alpha = .15, color = NA) +
  geom_line(linewidth = .7) + geom_point(size = 1.3) +
  scale_color_manual(values = c("grey55", COL_ACT, COL_PRED)) +
  scale_fill_manual(values = c("grey55", COL_ACT, COL_PRED)) +
  scale_x_log10(breaks = c(10, 40, 160), expand = expansion(mult = c(.10, .14))) +
  coord_cartesian(ylim = c(-.2, .95)) +
  labs(title = "E", x = "Codons sampled", y = expression("Variance explained ("*R^2*")")) +
  theme_fig() + guides(color = guide_legend(ncol = 1), fill = guide_legend(ncol = 1)) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom", legend.title = element_blank(),
        legend.text = element_text(size = 6, margin = margin(r = 2)), legend.key.size = unit(.3, "lines"),
        legend.box.margin = margin(-4, 0, 0, -6), plot.margin = margin(2, 2, 2, 2))

cd_block <- arrangeGrob(panel_C, panel_D, ncol = 1)
bottom   <- arrangeGrob(cd_block, panel_E, ncol = 2, widths = c(.75, .25))
HTS <- unit(c(2.15, 2.55, 4.05), "in")
figure7    <- arrangeGrob(panel_A, panel_B,    bottom, ncol = 1, heights = HTS)
figure7_nb <- arrangeGrob(panel_A, panel_B_nb, bottom, ncol = 1, heights = HTS)
for (ext in c("png", "pdf")) {
  ggsave(paste0("Figure7_main.", ext), figure7, width = 6.5, height = 8.75, dpi = 300, bg = "white")
  ggsave(paste0("Figure7_main_noBorderB.", ext), figure7_nb, width = 6.5, height = 8.75, dpi = 300, bg = "white")
}
grid.newpage(); grid.draw(figure7)
```

# 11b. Cross-gene transfer: LMNA-trained model applied to BRCA1

The six positional features are computable in any transcript, so transfer is directly testable:
train on LMNA only and predict BRCA1 without any BRCA1 training. BRCA1's readout (relative RNA
abundance) is inverted in sign and on a different scale from NMD efficiency, so the comparable
statistic is |Spearman|. The transferred model uses the positional features alone, so that the
test is confined to positional structure; adding stop identity and the +4/+5 context leaves
|rho| unchanged at 0.61.

```{r transfer, cache=TRUE}
sp <- function(a, b) suppressWarnings(cor(a, b, method = "spearman"))
fit_pos_only <- function(X, yy, idx)
  ranger(yy ~ ., data = cbind(X[idx, , drop = FALSE], yy = yy[idx]),
         num.trees = 500, min.node.size = 3, seed = SEED)
tr_B <- predict(fit_pos_only(Fpos, y, seq_len(n)), data = bF)$predictions
rho_model <- abs(sp(tr_B, by)); rho_nmdA <- abs(sp(bb$nmdA, by))
# paired bootstrap over codons, the independent unit in BRCA1
bidx <- split(seq_along(bc), bc); bucn <- names(bidx)
set.seed(SEED)
gapv <- replicate(2000, { i <- unlist(bidx[sample(bucn, length(bucn), replace = TRUE)], use.names = FALSE)
  abs(sp(tr_B[i], by[i])) - abs(sp(bb$nmdA[i], by[i])) })
transfer <- data.frame(direction = "LMNA-trained -> BRCA1",
  rho_model = rho_model, rho_nmdA = rho_nmdA,
  gap_median = median(gapv, na.rm = TRUE),
  gap_lo = quantile(gapv, .025, na.rm = TRUE), gap_hi = quantile(gapv, .975, na.rm = TRUE),
  p_better = mean(gapv > 0, na.rm = TRUE), row.names = NULL)
print(transfer, digits = 3, row.names = FALSE)
cat("  gap_* is |rho| of the transferred model minus |rho| of NMDetectiveA on the same BRCA1 codons,\n")
cat("  from a paired bootstrap over codons. The interval includes zero.\n")
write.csv(transfer, file.path(OUT, "cross_gene_transfer.csv"), row.names = FALSE)
```

# 12. Figure S7

```{r figureS7, fig.width=10, fig.height=13.0}
## A: classification cutoff
kde <- density(y[is_train], bw = .05, n = 2048, from = -.2, to = 1)
gdf <- data.frame(x = grid, c1 = w[1] * dnorm(grid, mu[1], sg[1]), c2 = w[2] * dnorm(grid, mu[2], sg[2]))
pA <- ggplot() + geom_area(data = data.frame(x = kde$x, yv = kde$y), aes(x, yv), fill = "#B7B7B7", alpha = .45) +
  geom_line(data = gdf, aes(x, c1), color = "#5B8FD1", linewidth = .9) +
  geom_line(data = gdf, aes(x, c2), color = COL_ACT, linewidth = .9) +
  geom_vline(xintercept = THRESH, color = "#C0392B", linetype = "dotted", linewidth = .9) +
  annotate("text", x = THRESH - .03, y = max(kde$y) * .6, label = sprintf("Cutoff %.2f", THRESH),
           color = "#C0392B", size = 2.9, hjust = 1) +
  annotate("text", x = .05, y = max(kde$y) * .30, label = "NMD-\nescaping", color = "#5B8FD1",
           size = 2.5, hjust = 0, lineheight = .9) +
  annotate("text", x = .88, y = max(kde$y) * .75, label = "NMD-\ntriggering", color = COL_ACT,
           size = 2.5, hjust = 1, lineheight = .9) +
  labs(title = "A", x = "Measured NMD efficiency", y = "Density") + theme_fig()

## B: SHAP beeswarm
clean <- function(v) { v <- sub("^nt_from_CDS_start$", "nt from CDS start", v)
  v <- sub("^nt_to_last_EJC$", "nt to last junction", v); v <- sub("^nt_to_next_EJC$", "nt to next junction", v)
  v <- sub("^nt_from_prev_EJC$", "nt from previous junction", v)
  v <- sub("^n_downstream_EJCs$", "no. downstream junctions", v)
  v <- sub("^exon_length$", "length of containing exon", v); v <- sub("^TGACT$", "TGA-CT context", v)
  v <- sub("^stop_", "stop = ", v); v <- sub("^plus4_", "+4 = ", v); v <- sub("^plus5_", "+5 = ", v); v }
Xt <- X_full[tr, ]; lv <- clean(names(imp))
shap_long <- do.call(rbind, lapply(names(imp), function(f) {
  val <- Xt[[f]]; rg <- range(val)
  data.frame(feature = factor(clean(f), levels = rev(lv)), shap = shap_values[, f],
             vnorm = if (diff(rg) == 0) .5 else (val - rg[1]) / diff(rg)) }))
pB <- ggplot(shap_long, aes(shap, feature, colour = vnorm)) +
  geom_vline(xintercept = 0, colour = "grey60", linewidth = .35) +
  geom_quasirandom(orientation = "y", method = "quasirandom", size = .35, alpha = .85, width = .38) +
  scale_colour_gradientn(colours = c("#2F6355", COL_ACT, "#8E9273", COL_PRED, "#A4611F"),
    breaks = c(0, 1), labels = c("Low", "High"), name = "Feature value",
    guide = guide_colourbar(barwidth = .32, barheight = 5.2, ticks = FALSE, title.position = "right",
                            title.theme = element_text(angle = -90, size = 7, hjust = .5))) +
  labs(title = "B", x = "SHAP value (impact on predicted NMD efficiency)", y = NULL) + theme_fig() +
  theme(panel.grid.major.y = element_line(colour = "grey92", linewidth = .3, linetype = "dotted"),
        axis.text.y = element_text(size = 7), legend.position = "right",
        legend.text = element_text(size = 7), legend.margin = margin(0, 0, 0, -2))

## C: how far the model can reach
sw <- sweep
sw$block <- factor(sw$block, levels = c("1", "3", "5", "10", "20", "30", "50", "tile"),
                   labels = c("1", "3", "5", "10", "20", "30", "50", "whole\nregion"))
swl <- rbind(data.frame(block = sw$block, model = "Our model", R2 = sw$R2),
             data.frame(block = sw$block, model = "NMDetectiveA (fixed reference)", R2 = NMDA_REF))
swl$model <- factor(swl$model, levels = c("Our model", "NMDetectiveA (fixed reference)"))
pC <- ggplot(swl, aes(block, R2, group = model, colour = model, linetype = model)) +
  geom_hline(yintercept = 0, colour = "grey70", linewidth = .3) +
  geom_errorbar(data = sw, inherit.aes = FALSE, aes(x = block, ymin = R2_lo, ymax = R2_hi),
                width = .2, colour = COL_ACT) +
  geom_line(linewidth = .6) +
  geom_point(data = swl[swl$model == "Our model", ], size = 1.6, show.legend = FALSE) +
  scale_colour_manual(values = c("Our model" = COL_ACT,
                                 "NMDetectiveA (fixed reference)" = "grey55")) +
  scale_linetype_manual(values = c("Our model" = "solid",
                                   "NMDetectiveA (fixed reference)" = "dashed")) +
  labs(title = "C", x = "Size of the hidden stretch (codons)",
       y = expression("Variance explained ("*R^2*")")) + theme_fig() +
  guides(colour = guide_legend(ncol = 1), linetype = guide_legend(ncol = 1)) +
  theme(axis.text.x = element_text(size = 7), legend.position = "bottom",
        legend.title = element_blank(), legend.text = element_text(size = 6),
        legend.key.size = unit(.55, "lines"), legend.box.margin = margin(-6, 0, 0, 0))

## D: the two transitions
# Fitted on individual PTCs rather than codon means, matching Fig. 7C,D. The
# midpoints are unchanged to two decimals: least squares on equal-sized group
# means and on all members share a minimiser, and almost every codon has 3 PTCs.
pcp <- ptc[!ptc$test, ]
logi <- function(x, p) p[1] + (p[2] - p[1]) / (1 + exp(-(x - p[3]) / p[4]))
bump <- function(x, p) (p[1] + p[2] * exp(-(x - p[3])^2 / (2 * p[4]^2))) / (1 + exp((x - p[5]) / p[6]))
fit_ls <- function(fn, dd, st, lo, hi) optim(st, function(p) sum((dd$actual - fn(dd$codon, p))^2),
                                             method = "L-BFGS-B", lower = lo, upper = hi)$par
p5 <- fit_ls(logi, pcp[pcp$codon <= 118, ], c(0, .85, 23, 1.5), c(-.2, .5, 10, .1), c(.3, 1, 40, 10))
p3 <- fit_ls(bump, pcp[pcp$codon >= 626, ], c(.1, .7, 633, 4, 640, 2), c(0, 0, 626, .5, 630, .2), c(1, 1, 645, 15, 650, 5))
z5 <- data.frame(codon = seq(1, 60, .25)); z5$fit <- logi(z5$codon, p5)
z3 <- data.frame(codon = seq(626, 656, .25)); z3$fit <- bump(z3$codon, p3)
zoom <- function(lo, hi, zf, ttl, leg) ggplot(ptc[ptc$codon >= lo & ptc$codon <= hi, ], aes(codon)) +
  geom_point(aes(y = actual, color = "Measured", shape = test), size = 1.25, alpha = .5) +
  geom_step(aes(y = ours, color = "Our model", group = stop), direction = "mid", linewidth = .5, alpha = .75) +
  geom_line(data = zf, aes(y = fit, color = "Fitted transition"), linetype = "dashed") +
  scale_color_manual(values = c(Measured = COL_ACT, "Our model" = COL_PRED, "Fitted transition" = "black"), name = NULL) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 1), guide = "none") +
  labs(title = ttl, x = "Codon", y = "NMD efficiency") + theme_fig() +
  guides(color = guide_legend(override.aes = list(alpha = 1))) +
  theme(legend.position = if (leg) "top" else "none", legend.key.size = unit(.4, "lines"),
        legend.text = element_text(size = 7, margin = margin(r = 5)), legend.box.margin = margin(0, 0, -5, 0))
pD1 <- zoom(1, 60, z5, "D", TRUE); pD2 <- zoom(626, 656, z3, "", FALSE)

## E: spread across the 50 held-out sets
show <- c("VEP predictor", "NMDetectiveA (RF)", "Our model")
rp <- rep_tab[rep_tab$model %in% show, ]; rp$model <- factor(rp$model, levels = show)
rl <- rbind(data.frame(model = rp$model, metric = "Codon ranking (AUC)", value = rp$AUC),
            data.frame(model = rp$model, metric = "Variance explained (R2)", value = rp$R2))
rl <- rl[!is.na(rl$value), ]
pE <- ggplot(rl, aes(model, value)) + geom_boxplot(aes(fill = model), width = .6, outlier.size = .5, linewidth = .3) +
  facet_wrap(~metric, scales = "free_y") +
  scale_fill_manual(values = c("grey80", "grey60", COL_ACT), guide = "none") +
  scale_x_discrete(labels = c("VEP", "NMDet.A", "Ours")) +
  labs(title = "E", x = NULL, y = "Value on held-out codons") + theme_fig() +
  theme(axis.text.x = element_text(size = 7))

## F: classification vs sample size
pF <- ggplot(lcp, aes(X, AUC_median, color = model, fill = model)) +
  geom_hline(yintercept = lcp$ref_AUC[1], linetype = "dashed", color = "grey35", linewidth = .4) +
  geom_ribbon(aes(ymin = AUC_q05, ymax = AUC_q95), alpha = .15, color = NA) +
  geom_line(linewidth = .7) + geom_point(size = 1.3) +
  scale_color_manual(values = c("grey55", COL_ACT, COL_PRED)) +
  scale_fill_manual(values = c("grey55", COL_ACT, COL_PRED)) +
  scale_x_log10(breaks = c(5, 10, 20, 40, 80, 160)) + coord_cartesian(ylim = c(.4, 1)) +
  labs(title = "F", x = "Codons sampled (1 PTC each)", y = "Codon ranking (AUC)") + theme_fig() +
  guides(color = guide_legend(ncol = 1), fill = guide_legend(ncol = 1)) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom", legend.title = element_blank(),
        legend.text = element_text(size = 7), legend.key.size = unit(.35, "lines"),
        legend.box.margin = margin(-4, 0, 0, 0))

## G: edit-reachable sampling
pG <- ggplot(sd, aes(X, R2_median, color = mode, fill = mode)) +
  geom_hline(yintercept = LC_REF_R2, linetype = "dashed", color = "grey35", linewidth = .4) +
  geom_ribbon(aes(ymin = R2_q05, ymax = R2_q95), alpha = .15, color = NA) +
  geom_line(linewidth = .7) + geom_point(size = 1.3) +
  scale_color_manual(values = c(COL_PRED, COL_EDIT), labels = c("Any codon", "Single-edit reachable")) +
  scale_fill_manual(values = c(COL_PRED, COL_EDIT), labels = c("Any codon", "Single-edit reachable")) +
  scale_x_log10(breaks = c(5, 10, 20, 30, 40, 60, 80)) + coord_cartesian(ylim = c(-.2, .95)) +
  labs(title = "G", x = "Codons sampled (1 PTC each)", y = expression("Variance explained ("*R^2*")")) +
  theme_fig() + guides(color = guide_legend(ncol = 1), fill = guide_legend(ncol = 1)) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom", legend.title = element_blank(),
        legend.text = element_text(size = 7), legend.key.size = unit(.35, "lines"),
        legend.box.margin = margin(-4, 0, 0, 0))

## H: BRCA1
bs <- bsum
bs$model <- factor(bs$model, levels = c("NMDetectiveA, calibrated", "Gene-specific model",
                                        "Gene-specific + NMDetectiveA"))
pH <- ggplot(bs, aes(X, R2_median, color = model, fill = model)) +
  geom_ribbon(aes(ymin = R2_q05, ymax = R2_q95), alpha = .15, color = NA) +
  geom_line(linewidth = .7) + geom_point(size = 1.3) +
  scale_color_manual(values = c("grey55", COL_ACT, COL_PRED)) +
  scale_fill_manual(values = c("grey55", COL_ACT, COL_PRED)) +
  scale_x_log10(breaks = c(10, 20, 30, 50)) + coord_cartesian(ylim = c(-.2, .95)) +
  labs(title = "H", x = "BRCA1 codons sampled", y = expression("Variance explained ("*R^2*")")) +
  theme_fig() + guides(color = guide_legend(ncol = 1), fill = guide_legend(ncol = 1)) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom", legend.title = element_blank(),
        legend.text = element_text(size = 7), legend.key.size = unit(.35, "lines"),
        legend.box.margin = margin(-4, 0, 0, 0))


## I, J: cross-gene transfer scatters
tdf <- rbind(data.frame(score = bb$nmdA, yv = by, who = "I"),
             data.frame(score = tr_B,    yv = by, who = "J"))
scat <- function(w, col, ttl, xlab, rho)
  ggplot(subset(tdf, who == w), aes(score, yv)) +
    geom_point(size = .9, colour = col, alpha = .8) +
    geom_smooth(method = "loess", formula = y ~ x, span = 1.6, se = FALSE,
                colour = "grey25", linewidth = .6) +
    annotate("text", x = Inf, y = Inf, hjust = 1.08, vjust = 1.6, size = 3, colour = col,
             fontface = "bold", label = sprintf("|\u03c1| = %.2f", rho)) +
    labs(title = ttl, x = xlab, y = "BRCA1 RNA score") +
    theme_fig() + theme(panel.grid.minor = element_blank())
pI <- scat("I", "grey45", "I", "NMDetectiveA score", rho_nmdA)
pJ <- scat("J", COL_ACT,  "J", "Predicted NMD efficiency (LMNA-trained)", rho_model)

figureS7 <- arrangeGrob(
  arrangeGrob(pA, pB, pC, ncol = 3, widths = c(.30, .40, .30)),
  arrangeGrob(pD1, pD2, pE, ncol = 3, widths = c(.38, .28, .34)),
  arrangeGrob(pF, pG, pH, ncol = 3, widths = c(.33, .33, .34)),
  arrangeGrob(pI, pJ, grid::nullGrob(), ncol = 3, widths = c(.33, .33, .34)),
  ncol = 1, heights = c(1, 1, 1, .85))
for (ext in c("png", "pdf")) ggsave(paste0("FigureS7_supplementary.", ext), figureS7,
                                    width = 10, height = 13.0, dpi = 300, bg = "white")
grid.newpage(); grid.draw(figureS7)
```

# 13. Table S7

```{r tableS7}
tabS7 <- data.frame(
  codon = lmna$codon, mutation = lmna$mutation,
  mean_NMD_efficiency = lmna$mean_NMD_efficiency, sd_NMD_efficiency = lmna$sd_NMD_efficiency,
  region = c("5' inactive", "active", "3' inactive")[1 + (codon >= 25) + (codon > 635)],
  plus4 = lmna$plus4, plus5 = lmna$plus5,
  X_full[, c("nt_from_CDS_start", "nt_to_last_EJC", "nt_to_next_EJC",
             "nt_from_prev_EJC", "exon_length", "n_downstream_EJCs")],
  TGACT = X_full$TGACT,
  rule5055_pred = rule_5055, aenmd_pred = rule_aenmd, vep_pred = rule_vep,
  NMDetectiveA = lmna$NMDetectiveA, NMDetectiveB = lmna$NMDetectiveB,
  true_label = lab, heldout_set = ifelse(is_test, "held-out", "training"),
  rf_pred = pred_plot, stringsAsFactors = FALSE)
keyS7 <- data.frame(Column = names(tabS7), Description = c(
 "Codon position in the LMNA coding sequence (NM_170707.4)",
 "Stop-codon identity introduced at this position (TAA, TAG, or TGA)",
 "Measured NMD efficiency, defined as 1 - (DMSO/SMG1i), averaged across replicate SNVs",
 "Standard deviation of NMD efficiency across replicate SNVs",
 "5' inactive / active / 3' inactive assignment. Descriptive annotation only; NOT a model input",
 "Nucleotide at the +4 position, immediately downstream of the PTC",
 "Nucleotide at the +5 position",
 "Model feature: nucleotides from the CDS start to the PTC",
 "Model feature: nucleotides from the PTC to the last exon-exon junction",
 "Model feature: nucleotides to the nearest downstream exon-exon junction (-1 if none)",
 "Model feature: nucleotides from the nearest upstream exon-exon junction",
 "Model feature: length of the exon containing the PTC, including untranslated sequence for exon 1",
 "Model feature: number of exon-exon junctions downstream of the PTC",
 "Model feature: 1 if a TGA stop is followed by C at +4 and T at +5, else 0",
 "Canonical 50-nt rule call (1 = NMD-triggering, 0 = NMD-escaping)",
 "aenmd rule-set call (1 = NMD-triggering, 0 = NMD-escaping)",
 "VEP NMD predictor call (1 = NMD-triggering, 0 = NMD-escaping)",
 "NMDetectiveA score, published per nucleotide position and therefore identical for the three stop codons at a codon",
 "NMDetectiveB score, published per nucleotide position and therefore identical for the three stop codons at a codon",
 sprintf("Ground-truth class from the Gaussian-mixture threshold of %.3f (1 = NMD-triggering, 0 = NMD-escaping)", THRESH),
 "Training or held-out portion of the prespecified split used for Figs. 7B-D",
 "Random forest prediction: codon-grouped out-of-fold for training codons, final-model for held-out codons"),
 stringsAsFactors = FALSE)
stopifnot(nrow(keyS7) == ncol(tabS7))
wbS7 <- createWorkbook()
addWorksheet(wbS7, "TableS7_NMD-modeling"); writeData(wbS7, 1, tabS7)
addWorksheet(wbS7, "Key"); writeData(wbS7, 2, keyS7)
setColWidths(wbS7, 2, cols = 1:2, widths = c(29, 117))
modifyBaseFont(wbS7, fontSize = 12, fontName = "Aptos Narrow")
saveWorkbook(wbS7, file.path(OUT, "TableS7_NMD-modeling.xlsx"), overwrite = TRUE)
cat(sprintf("Table S7 written: %d rows x %d columns\n", nrow(tabS7), ncol(tabS7)))
```

# 14. Every number quoted in the manuscript

```{r numbers-for-text}
say <- function(...) cat(sprintf(...), "\n")
say("Misclassified, PTC level (of %d): %s", nrow(ptc),
    paste(sprintf("%s %d", names(n_ptc), n_ptc), collapse = " | "))
say("Misclassified, codon level (of %d): %s", nrow(pc),
    paste(sprintf("%s %d", names(n_cod), n_cod), collapse = " | "))
say("AUC medians:  %s", paste(sprintf("%s %.2f", rs[key, "model"], rs[key, "AUC_median"]), collapse = " | "))
say("R2 medians:   %s", paste(sprintf("%s %.2f", rs[key, "model"], rs[key, "R2_median"]), collapse = " | "))
say("Our model beats NMDetectiveA on R2 in %d of %d splits",
    sum(rep_tab$R2[rep_tab$model == "Our model"] > rep_tab$R2[rep_tab$model == "NMDetectiveA (RF)"]), N_SPLITS)
say("\n--- Where the errors fall (PTC level) ---")
w <- err_ptc[["Our model"]] != ptc$label
e <- ptc[w, c("codon", "stop")]; e$type <- ifelse(err_ptc[["Our model"]][w] == 1, "FP", "FN")
say("Our model misclassifies %d of %d PTCs:", sum(w), nrow(ptc))
print(e[order(e$codon), ], row.names = FALSE)
zone <- (e$codon >= 24 & e$codon <= 35) | (e$codon >= 630 & e$codon <= 645)
say("  in the 5' and 3' transition zones: %d of %d; elsewhere: %s",
    sum(zone), nrow(e), paste(sort(unique(e$codon[!zone])), collapse = ", "))
tga_only <- sapply(sort(unique(e$codon)), function(cc) {
  at <- err_ptc[["Our model"]][ptc$codon == cc] != ptc$label[ptc$codon == cc]
  sum(at) == 1 && as.character(ptc$stop[ptc$codon == cc][at]) == "TGA" })
say("  errors carried by TGA alone at that codon: %s",
    paste(sort(unique(e$codon))[tga_only], collapse = ", "))
first_exon <- ptc$codon <= 118
for (m in c("aenmd", "VEP", "NMDetectiveB")) say("%s: %d first-exon false negatives; %.0f%% of first-exon PTCs correct",
    m, sum(err_ptc[[m]] == 0 & ptc$label == 1 & first_exon),
    100 * mean(err_ptc[[m]][first_exon] == ptc$label[first_exon]))
fp5055 <- sort(unique(ptc$codon[err_ptc[["50-nt rule"]] == 1 & ptc$label == 0]))
say("50-nt rule false positives: %d codons, from %d to %d (not contiguous; %d below codon 100)",
    length(fp5055), min(fp5055), max(fp5055), sum(fp5055 < 100))
say("\n--- Discordance ---")
say("Discordance threshold %.3f per PTC (%.3f per codon mean)", DISCORD, DISCORD_CODON)
say("Discordant PTCs: ours %d of %d, NMDetectiveA %d of %d", 
    sum(abs(ptc$actual - ptc$ours) > DISCORD), nrow(ptc),
    sum(abs(ptc$actual - ptc$nmdA) > DISCORD), nrow(ptc))
say("Ours, discordant PTCs: %s", paste(sprintf("%d%s",
    ptc$codon[abs(ptc$actual - ptc$ours) > DISCORD],
    ptc$stop[abs(ptc$actual - ptc$ours) > DISCORD]), collapse = ", "))
say("\n--- Variance structure and the TGA-CT effect ---")
cm <- ave(y, codon, FUN = mean)
say("Variance: total %.4f | between codons %.1f%% | within codons %.1f%% (unreachable by a per-codon score)",
    var(y), 100 * var(cm) / var(y), 100 * var(y - cm) / var(y))
on <- X_full$TGACT[tr] == 1
say("TGA-CT present in %d of %d training PTCs; mean SHAP when present %+.3f (min %+.3f), when absent %+.3f",
    sum(on), length(on), mean(shap_values[on, "TGACT"]), min(shap_values[on, "TGACT"]), mean(shap_values[!on, "TGACT"]))
say("Codon 240: TAA %.2f, TAG %.2f, TGA %.2f (measured drop %.2f)",
    y[codon == 240 & lmna$mutation == "TAA"], y[codon == 240 & lmna$mutation == "TAG"],
    y[codon == 240 & lmna$mutation == "TGA"],
    y[codon == 240 & lmna$mutation == "TAA"] - y[codon == 240 & lmna$mutation == "TGA"])
say("Codon 217: %s", paste(sprintf("%s %.2f", lmna$mutation[codon == 217], y[codon == 217]), collapse = ", "))
say("Models miscalling at least one PTC at codon 217: %s",
    paste(names(err_ptc)[sapply(err_ptc, function(p) any(p[ptc$codon == 217] != ptc$label[ptc$codon == 217]))], collapse = ", "))
say("\n--- Sampling requirement ---")
g <- lc[lc$model == "Gene-specific RF", ]
say("NMDetectiveA calibrated on the whole pool: R2 %.3f, AUC %.3f", LC_REF_R2, LC_REF_AUC)
say("Gene-specific model: R2 %.2f at 20 sites, %.2f at 40, %.2f at 80",
    g$R2_median[g$X == 20], g$R2_median[g$X == 40], g$R2_median[g$X == 80])
say("Crossover: median exceeds the reference from X = %d; in every draw from X = %d", X_CROSS, X_ROBUST)
se <- sd[sd$mode == "Single-edit-reachable codons only", ]
say("Edit-reachable: %d of %d assayed codons; crossover at X = %d; R2 %.2f at 50",
    length(unique(codon[one_edit])), length(ucod), se$X[which(se$R2_median > LC_REF_R2)[1]],
    se$R2_median[se$X == 50])
say("BRCA1 at 10 sites: gene-specific %.2f vs NMDetectiveA calibrated %.2f",
    bsum$R2_median[bsum$X == 10 & bsum$model == "Gene-specific model"],
    bsum$R2_median[bsum$X == 10 & bsum$model == "NMDetectiveA, calibrated"])
say("Transfer LMNA -> BRCA1: |rho| model %.2f vs NMDetectiveA %.2f; gap %+.2f [%+.2f, %+.2f], better in %.0f%% of resamples",
    transfer$rho_model, transfer$rho_nmdA, transfer$gap_median, transfer$gap_lo,
    transfer$gap_hi, 100 * transfer$p_better)
say("\n--- Boundaries ---")
say("measured %s | NMDetectiveA %s | NMDetectiveB %s | ours %s", emp_act, emp_A, emp_B, emp_ours)
say("5' logistic midpoint: codon %.2f (%.0f nt) | 3' drop midpoint: codon %.2f (%.0f nt)",
    p5[3], 3 * (p5[3] - 1), p3[5], LAST_EJC - 3 * p3[5])
say("NMDetectiveA takes %d distinct values over the assayed codons; NMDetectiveB %d",
    length(unique(round(lmna$NMDetectiveA, 10))), length(unique(round(lmna$NMDetectiveB, 10))))
say("NMDetectiveB distinct values across codons 214-566: %d",
    length(unique(lmna$NMDetectiveB[codon >= 214 & codon <= 566])))
```

# Session info

```{r session}
sessionInfo()
```
